\textbf{ Normalized Gaussian Processes }

### 2020-11-30 tex/ms.tex
better notation

\section{The problem}

Let $\mathbbb{x} = \left( \mathbb{x}_0 \,\, \cdots \,\, \mathbb{x}_{K-1} \right)^\top$
be a $K$-dimensional multivariate normal random variable distributed
according to a Gaussian Process with mean $\pmb{\mu}$
and covariance $\pmb{\Sigma}$:
%
\begin{align}
    \mathbbb{x} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad.
\end{align}
%
For simplicity, let us assume the mean is constant, i.e.,
%
\begin{align}
    \pmb{\mu} = \mu \, \mathbf{j}
\end{align}
%
where $\mathbf{j}$ is the vector of $K$ ones.
%
Now suppose we cannot observe samples of $\mathbbb{x}$ directly, but instead we can
observe samples from the \emph{normalized} process, which we will call
$\btilde{\mathbbb{x}}$:
%
\begin{align}
    \btilde{\mathbbb{x}}
     & \equiv \frac{\mathbbb{x}}{\bbar{\mathbb{x}}}
    \nonumber                                                              \\[0.5em]
     & = \frac{\mathbbb{x}}{\frac{1}{K}\sum\limits_{k=0}^{K-1} \mathbb{x}}
    \quad.
\end{align}
%
The random variable $\btilde{\mathbbb{x}}$ is no longer normally distributed, but as long as
the variance of $\bbar{\mathbb{x}}$ is small, we can attempt to approximate it as such. Let the mean
and covariance of $\btilde{\mathbbb{x}}$ be $\btilde{\pmb{\mu}}$ and $\btilde{\pmb{\Sigma}}$, respectively:
%
\begin{align}
    \btilde{\mathbbb{x}} \mathrel{\dot\sim} \mathcal{N}\left( \btilde{\pmb{\mu}}, \btilde{\pmb{\Sigma}} \right)
    \quad.
\end{align}
%
By definition, the mean is unity, i.e,
%
\begin{align}
    \btilde{\pmb{\mu}} = \mathbf{j}
    \quad.
\end{align}
%
What is the expression for $\btilde{\pmb{\Sigma}}$?

\section{The solution}

The covariance is given by

@@

        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{j}\right)
        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{j}\right)^\top

@@

        \frac{\mathbbb{x}\,\mathbf{j}^\top}{\bbar{\mathbb{x}}}

@@

        \frac{\mathbf{j}\,\mathbbb{x}^\top}{\bbar{\mathbb{x}}}

@@

        \mathbf{j} \, \mathbf{j}^\top

@@

To evaluate this, it is convenient to write

@@

    \mathbbb{u} \sim \mathcal{N}\left( 0, \mathbf{I} \right)

@@

            (\pmb{\mu} + \mathbf{L} \mathbbb{u})\mathbf{j}^\top

@@

            \mathbf{j}(\pmb{\mu} + \mathbf{L} \mathbbb{u})^\top

@@

    \mu + \frac{1}{K}\mathbf{j}^\top \mathbf{L} \, \mathbbb{u}

@@

     & = \frac{1}{\mu K}\mathbf{j}^\top \mathbf{L} \, \mathbbb{u}

@@

        \frac{\mathbb{r}^2 \, \mathbf{j} \, \mathbf{j}^\top}{(1 + \mathbb{r})^2}

@@

        \frac{\mathbb{r}\, \mathbf{L} \, \mathbbb{u} \, \mathbf{j}^\top}{(1 + \mathbb{r})^2}

@@

        \mathbf{j}^\top

@@

        \mathbb{r}^{n + 2} \, \mathbf{j} \, \mathbf{j}^\top

@@

    \frac{(n + 1) g_{n} m^\frac{n+2}{2}}{\mu^{n+2}} \, \mathbf{j} \, \mathbf{j}^\top

@@

        \mathbb{r}^{n + 1} \, \mathbf{L} \, \mathbbb{u} \, \mathbf{j}^\top

@@

    \frac{(n + 1) K \, g_{n} m^\frac{n}{2}}{\mu^{n+1}} \, \mathbf{m} \, \mathbf{j}^\top

@@

    n \,  m^\frac{n - 2}{2} \mathbf{m} \, \mathbf{m}^\top + m^\frac{n}{2} \pmb{\Sigma}

@@

    \mathbf{m}

@@

    \frac{1}{K} \pmb{\Sigma} \, \mathbf{j}

@@

is the average of each row in $\pmb{\Sigma}$, and

@@

@@

    \frac{1}{K^2} \mathbf{j}^\top \pmb{\Sigma} \, \mathbf{j}

@@

is the average of all elements in $\pmb{\Sigma}$.
Inserting these expressions into Equation~(\ref{eq:SigmaPQR}) and rearranging, we obtain
the final expression for the normalized covariance,
    \label{eq:SigmaTilde}
    \frac{\alpha \, \pmb{\Sigma}}{\mu^2}  +
    \frac{\alpha \, \left(\mathbf{s} \, \mathbf{s}^\top - \mathbf{m} \, \mathbf{m}^\top\right)}{\mu^2 m}  +
    \frac{\beta \, \left(\mathbf{s} \, \mathbf{s}^\top\right)}{\mu^2 m}
where we define
    \mathbf{s}
    m \, \mathbf{j} - \mathbf{m}
    \\[1em]
    \label{eq:alpha}
    \alpha
    \frac{(2n + 1)!}{2^n \, n!}
    z^n
    \\[1em]
    \label{eq:beta}
    \beta
    \frac{2n(2n + 1)!}{2^n \, n!}
    z^n
as well as the dimensionless parameter
    z & \equiv \frac{m}{\mu^2}
equal to the average element of the covariance matrix divided by the square of the
GP mean.

\section{The Bad News}
%
In the previous section we derived a compact expression for
the covariance of a normalized Gaussian process (Equation~\ref{eq:SigmaTilde}),
which can be computed as the sum of a term proportional to the original
covariance $\pmb{\Sigma}$ and a low-rank correction.
%
Unfortunately, the expressions for the scaling constants $\alpha$ and
$\beta$ (Equations~\ref{eq:alpha} and \ref{eq:beta}) involve series
expansions that \emph{do not converge}.
%
\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/alpha_beta.pdf}
        \oscaption{alpha_beta}{%
            Asymptotic series expansions for $\alpha$ and $\beta$
            (Equations~\ref{eq:alpha} and \ref{eq:beta}).
            The left panel shows the value of each
            coefficient ($\alpha$: solid, $\beta$: dashed)
            as a function of the index $n$ for three
            different values of $z$. In all cases, the value initially decreases
            with $n$ but eventually diverges to $+\infty$ as $n \rightarrow \infty$.
            The vertical lines indicate the optimal truncation order $N$,
            for which the partial sums yield the best approximation to the
            asymptotic values of the cofficients. The right panels show the
            partial sums evaluated at each order; the value at the optimal
            truncation order is marked with the horizontal dashed lines.
            \label{fig:alpha_beta}
        }
    \end{centering}
\end{figure}
%
This is demonstrated in Figure~\ref{fig:alpha_beta}, which shows the
terms in the summations in Equations~\ref{eq:alpha} and \ref{eq:beta} for
different values of $z$. While the terms initially decrease in magnitude
(left panel), asymptotically approaching a finite value for both
$\alpha$ and $\beta$ (right panels), as the expansion order increases
the terms eventually diverge, leading to infinite values for both
coefficients. Even though the series take longer to diverge for smaller
values of $z$, the divergence occurs for all $z \ne 0$: i.e., the radius
of convergence for both series is zero.

@@

>>>>

\section{Normalized Gaussian Processes}
\label{sec:main}

\subsection{Notation}
\label{sec:notation}

We adopt
the following conventions throughout this paper:
scalars are represented by italic lowercase letters (i.e., $x$)
and occasionally upercase letters, in particular if the value
corresponds to an integer (i.e., $K$);
column vectors are
represented by boldface lowercase letters
($\mathbf{x}$); and matrices and other linear operators are represented
by boldface capital letters ($\mathbf{X}$).
To make the distinction between numerical values and random variables as
clear as possible, we use blackboard fonts to denote random variables
(i.e., $\mathbb{x}$ for a scalar random variable and $\mathbbb{x}$
for a vector-valued random variable).
Finally, we use a bar symbol
($\bbar{\mathbb{x}}$) to denote the mean of a vector-valued random
variable and the tilde symbol
($\btilde{\mathbf{x}}$ or $\btilde{\mathbbb{x}}$) to denote a numerical
quantity or random variable corresponding
to a normalized Gaussian process.

\subsection{The problem}
\label{sec:problem}

Let $\mathbbb{x} = \left( \mathbb{x}_0 \,\, \cdots \,\, \mathbb{x}_{K-1} \right)^\top$
be a $K$-dimensional multivariate normal random variable distributed
according to a Gaussian Process with mean $\pmb{\mu}$
and covariance $\pmb{\Sigma}$:
%
\begin{align}
    \mathbbb{x} \sim \mathcal{N}\left( \pmb{\mu}, \pmb{\Sigma} \right)
    \quad.
\end{align}
%
For simplicity, let us assume the mean is constant, i.e.,
%
\begin{align}
    \pmb{\mu} = \mu \, \mathbf{1}
\end{align}
%
where $\mathbf{1}$ is the vector of $K$ ones.
%
Now suppose we cannot observe samples of $\mathbbb{x}$ directly, but instead we can
observe samples from the \emph{normalized} process, which we will call
$\btilde{\mathbbb{x}}$:
%
\begin{align}
    \btilde{\mathbbb{x}}
     & \equiv \frac{\mathbbb{x}}{\bbar{\mathbb{x}}}
    \nonumber                                                              \\[0.5em]
     & = \frac{\mathbbb{x}}{\frac{1}{K}\sum\limits_{k=0}^{K-1} \mathbb{x}}
    \quad.
\end{align}
%
The random variable $\btilde{\mathbbb{x}}$ is no longer normally distributed, but as long as
the variance of $\bbar{\mathbb{x}}$ is small, we can attempt to approximate it as such. Let the mean
and covariance of $\btilde{\mathbbb{x}}$ be $\btilde{\pmb{\mu}}$ and $\btilde{\pmb{\Sigma}}$, respectively:
%
\begin{align}
    \btilde{\mathbbb{x}} \mathrel{\dot\sim} \mathcal{N}\left( \btilde{\pmb{\mu}}, \btilde{\pmb{\Sigma}} \right)
    \quad.
\end{align}
%
By definition, the mean is unity, i.e,
%
\begin{align}
    \btilde{\pmb{\mu}} = \mathbf{1}
    \quad.
\end{align}
%
What is the expression for $\btilde{\pmb{\Sigma}}$?

\subsection{The solution}
\label{sec:solution}

From the definition of the covariance, we may compute $\btilde{\pmb{\Sigma}}$ from

@@

        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{1}\right)
        \left(\frac{\mathbbb{x}}{\bbar{\mathbb{x}}} - \mathbf{1}\right)^\top

@@

        \frac{\mathbbb{x}\,\mathbf{1}^\top}{\bbar{\mathbb{x}}}

@@

        \frac{\mathbf{1}\,\mathbbb{x}^\top}{\bbar{\mathbb{x}}}

@@

        \mathbf{1} \, \mathbf{1}^\top

@@

where $\mathbbb{E}\left[ \cdots \right]$ denotes the expected value of a random variable.
To evaluate the expression above, it is convenient to write

@@

    \mathbbb{u} \sim \mathcal{N}\left( \mathbf{0}, \mathbf{I} \right)

@@

            (\pmb{\mu} + \mathbf{L} \mathbbb{u})\mathbf{1}^\top

@@

            \mathbf{1}(\pmb{\mu} + \mathbf{L} \mathbbb{u})^\top

@@

    \mu + \frac{1}{K}\mathbf{1}^\top \mathbf{L} \, \mathbbb{u}

@@

     & = \frac{1}{\mu K}\mathbf{1}^\top \mathbf{L} \, \mathbbb{u}

@@

        \frac{\mathbb{r}^2 \, \mathbf{1} \, \mathbf{1}^\top}{(1 + \mathbb{r})^2}

@@

        \frac{\mathbb{r}\, \mathbf{L} \, \mathbbb{u} \, \mathbf{1}^\top}{(1 + \mathbb{r})^2}

@@

        \mathbf{1}^\top

@@

        \mathbb{r}^{n + 2} \, \mathbf{1} \, \mathbf{1}^\top

@@

    \frac{(n + 1) g_{n} m^\frac{n+2}{2}}{\mu^{n+2}} \, \mathbf{1} \, \mathbf{1}^\top

@@

        \mathbb{r}^{n + 1} \, \mathbf{L} \, \mathbbb{u} \, \mathbf{1}^\top

@@

    \frac{(n + 1) K \, g_{n} m^\frac{n+2}{2}}{\mu^{n+1}} \, \mathbf{q} \, \mathbf{1}^\top

@@

    n \,  m^\frac{n + 2}{2} \mathbf{q} \, \mathbf{q}^\top + m^\frac{n}{2} \pmb{\Sigma}

@@

@@

    \frac{1}{K^2} \mathbf{1}^\top \pmb{\Sigma} \, \mathbf{1}

@@

is the average of all elements in $\pmb{\Sigma}$, and

@@

    \mathbf{q}

@@

    \frac{1}{K m} \pmb{\Sigma} \, \mathbf{1}

@@

is the ratio of the average of each row in $\pmb{\Sigma}$ to the average of
all elements in $\pmb{\Sigma}$.

%
Inserting these expressions into Equation~(\ref{eq:SigmaPQR}) and rearranging, we obtain
the final expression for the normalized covariance,
%
\begin{align}
    \label{eq:SigmaTilde}
    \btilde{\pmb{\Sigma}}
     & =
    \frac{\alpha}{\mu^2} \pmb{\Sigma} +
    z \Big(
    (\alpha + \beta) \, \mathbf{p} \, \mathbf{p}^\top
    - \alpha \, \mathbf{q} \, \mathbf{q}^\top
    \Big)
    \quad,
\end{align}
%
where we define
%
\begin{align}
    \mathbf{p}
     & \equiv
    \mathbf{1} - \mathbf{q}
    \\[1em]
    \label{eq:alpha}
    \alpha
     & \equiv
    \sum\limits_{n=0}^\infty
    \frac{(2n + 1)!}{2^n \, n!}
    z^n
    \\[1em]
    \label{eq:beta}
    \beta
     & \equiv
    \sum\limits_{n=0}^\infty
    \frac{2n(2n + 1)!}{2^n \, n!}
    z^n
\end{align}
%
as well as the dimensionless parameter
%
\begin{align}
    z & \equiv \frac{m}{\mu^2}
    \quad,
\end{align}
%
equal to the average element of the covariance matrix divided by the square of the
GP mean.

\subsection{The Bad News}
\label{sec:badnews}
%
In the previous section we derived a compact expression for
the covariance of a normalized Gaussian process (Equation~\ref{eq:SigmaTilde}),
which can be computed as the sum of a term proportional to the original
covariance $\pmb{\Sigma}$ and a low-rank correction.
%
Unfortunately, the expressions for the scaling constants $\alpha$ and
$\beta$ (Equations~\ref{eq:alpha} and \ref{eq:beta}) involve series
expansions that \emph{do not converge}.
%
\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/alpha_beta.pdf}
        \oscaption{alpha_beta}{%
            Asymptotic series expansions for $\alpha$ and $\beta$
            (Equations~\ref{eq:alpha} and \ref{eq:beta}).
            The left panel shows the value of each
            coefficient ($\alpha$: solid, $\beta$: dashed)
            as a function of the index $n$ for three
            different values of $z$. In all cases, the value initially decreases
            with $n$ but eventually diverges to $+\infty$ as $n \rightarrow \infty$.
            The vertical lines indicate the optimal truncation order $N$,
            for which the partial sums yield the best approximation to the
            asymptotic values of the cofficients. The right panels show the
            partial sums evaluated at each order; the value at the optimal
            truncation order is marked with the horizontal dashed lines.
            \label{fig:alpha_beta}
        }
    \end{centering}
\end{figure}
%
This is demonstrated in Figure~\ref{fig:alpha_beta}, which shows the
terms in the summations in Equations~\ref{eq:alpha} and \ref{eq:beta} for
different values of $z$. While the terms initially decrease in magnitude
(left panel), asymptotically approaching a finite value for both
$\alpha$ and $\beta$ (right panels), as the expansion order increases
the terms eventually diverge, leading to infinite values for both
coefficients. Even though the series take longer to diverge for smaller
values of $z$, the divergence occurs for all $z \ne 0$: i.e., the radius
of convergence for both series is zero.

@@

%
\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/alpha_beta_values.pdf}
        \oscaption{alpha_beta_values}{%
            Caption.
            \label{fig:alpha_beta_values}
        }
    \end{centering}
\end{figure}
%
